### 2016-03-17 Poster/main.tex
tabelle eingeführt, neue graphen, lupen, mehr referenzen

%\usepackage{wrapfig}
%\usepackage{fancyhdr}
\usepackage{float}
 %\usepackage{graphicx}
\usepackage{subfig}

@@

\newlength{\sepwid}
\newlength{\onecolwid}
\newlength{\twocolwid}
\newlength{\threecolwid}
\setlength{\paperwidth}{1189mm} % A0 width: 46.8in
\setlength{\paperheight}{841mm} % A0 height: 33.1in
\setlength{\sepwid}{0.024\paperwidth} % Separation width (white space) between columns
\setlength{\onecolwid}{0.22\paperwidth} % Width of one column
\setlength{\twocolwid}{0.464\paperwidth} % Width of two columns
\setlength{\threecolwid}{0.708\paperwidth} % Width of three columns
\setlength{\topmargin}{-0.5in} % Reduce the top margin size
%-----------------------------------------------------------

@@

@@

@@

@@

We analysed the storage system of a super computer and tried to find a good predictor for its I/O-performance using artificial neural networks (ANNs).
A hypothesis was then made: \textbf{The I/O-path has significant influence on the time needed for a file access.}
In particular the access time for a file access is more dependent on the I/O-path used in the storage system than on the access size for example.\medskip

To test our hypothesis we used ANNs with different input information for the prediction of access times.
For the use of I/O-paths as input information for the ANNs we developed a method to compute an approximation of the different I/O-paths used in an benchmark-test.

@@

	Our measurements showed, as can be seen in figure \ref{exploration}, that the access times increase with access sizes.
	However a phenomenon that can be explained with the I/O-path model can be seen as well; \textbf{access times of measurements with equal parameter values are split into several groups of access times.} Different I/O-paths were used for file accesses in these groups, which led to a step in the magnitude of the access time.

	\begin{figure}
		\subfloat[Sequential access pattern]{
			\includegraphics[width=.58\textwidth]{src/plot_SizeSorted_log_read_seq.png}
		}
		%\hfill
		\\
		\subfloat[Random access pattern]{
			\includegraphics[width=.58\textwidth]{src/plot_SizeSorted_log_read_rnd.png}
		}	
		\caption{Access times of measured file accesses. The access sizes increase from left to right, all measurements with equal parameter values have the same color.}
		\label{exploration}
	\end{figure}

@@

	Different models were used for the prediction of measured access times of file accesses.
	As baseline model linear regression was used for a simple mapping of access size to measured access time.
	Additionally three \textbf{models with different input information utilizing ANNs were used}.
	Every ANN-model received information about access sizes, file offsets and access types of an input.
	One ANN also received information about the past data throughputs of the system, this can be used to exploit time dependencies of the I/O-performance in the system (see figure \ref{time_dep}). The last ANN received error classes as additional input.

	\begin{figure}
		\subfloat[In this sequence peaks in the access times occure regularly.]{
			\includegraphics[width=.48\textwidth]{src/plot_From100001to100250_read_seq.png}
		}
		\hfill
		\subfloat[Exponential moving average in red to predict peaks.]{
			\includegraphics[width=.48\textwidth]{src/ema2.png}
		}	
		\caption{Time dependencies can be utilized for better access time predictions.}
		\label{time_dep}
	\end{figure}

@@

The idea of error classes is based on the fact that \textbf{the residues of a simple model are characteristic for the groups of measurements with different typical access times} and therefore also for the I/O-path used by the storage system for the corresponding file accesses. 
%Here a simple model is a model that doesn't discriminate measurements with similar parameters.
Error classes are then obtained by clustering the residues of such a model with a k-Means algorithm.

	\begin{figure}
		\subfloat[Sequential access pattern]{
			\includegraphics[width=.58\textwidth]{src/linreg_error_clustering_seq_all.png}
		}
		\\
		%\hfill
		\subfloat[Random access pattern]{
			\includegraphics[width=.58\textwidth]{src/linreg_error_clustering_rnd_all.png}
		}	
		\caption{Residues from the linear regression model clustered into error classes.}
		\label{error_classes}
	\end{figure}

@@

	From the residues of the different models (see table \ref{residues}) it became clear that access times shouldn't be modelled linearly.
	Furthermore the exploitation of time dependencies of the storage system has been found to be difficult.
	As our hypothesis predicted \textbf{it is essential to use knowledge about I/O-paths for precise predictions of access times}.
	In figure \ref{preds} the difference in the predictions of the ANN without and with error class input can be seen.
	The second ANN is able to classify file accesses correctly in respect to their I/O-path.

	\begin{table}
		\vspace{2ex}
		\begin{tabular}{l|r|r}%
			&  \multicolumn{2}{|c}{mean absolut error}\\ \hline
			%\toprule
			Modell & seq. access & rnd. access\\ \hline
			%\midrule
			Linear regression & $7.6\cdot 10^-5$ & $4.76\cdot 10^-3$ \\
			ANN & $6.0\cdot 10^-5$ & $3.13\cdot 10^-3$ \\
			ANN + time dependencies & $5.7\cdot 10^-5$ & $3.05\cdot 10^-3$\\
			ANN + error classes & $2.0\cdot 10^-5$ & $1.03\cdot 10^-3$ \\
			%\bottomrule
		\end{tabular}
		\caption{Mean absolut errors of the different models for the two different file access patterns.}
		\label{residues}
	\end{table}

@@

	\vspace*{-2cm}
	\begin{figure}
		\subfloat[ANN without error classes]{
			\includegraphics[width=.47\textwidth]{src/plot_onlyPred_tuple1_Duration_rnd.png}
		}
		\hfill		
		\subfloat[ANN with error classes]{
			\includegraphics[width=.47\textwidth]{src/plot_onlyPred_tuple1_with_error_class_from_linreg_Duration_rnd.png}
		}	
		\caption{Measurements with random access pattern. Measured access times in black; predictions in blue.}
		\label{preds}
	\end{figure} 
%	
%	\begin{figure}
%		\includegraphics[width=0.48\linewidth]{src/plot_onlyPred_tuple1_Duration_rnd.png}
%		\hfill
%		\includegraphics[width=0.48\linewidth]{src/plot_onlyPred_tuple1_with_error_class_from_linreg_Duration_rnd.png}
%		\vspace*{-0.5cm}
%		\caption{Measurements in black; predictions in blue.}
%		\label{preds}
%	\end{figure}

@@

\nocite{*} % Insert publications even if they are not cited in the poster
\footnotesize{\bibliographystyle{unsrt}
\bibliography{sample}\vspace{0.75in}}

@@

@@

%----------------------------------------------------------------------------------------
%	CONTACT INFORMATION
%----------------------------------------------------------------------------------------

\setbeamercolor{block alerted title}{fg=black,bg=Apricot} % Change the alert block title colors
\setbeamercolor{block alerted body}{fg=black,bg=white} % Change the alert block body colors
\vspace*{-1.5cm}
\begin{alertblock}{Contact Information}

\begin{itemize}
%\item Web: \href{http://wr.informatik.uni-hamburg.de}{http://wr.informatik.uni-hamburg.de}
\item \href{mailto:2schmid@informatik.uni-hamburg.de}{2schmid@informatik.uni-hamburg.de}
\item \href{mailto:juliankunkel@googlemail.com}{juliankunkel@googlemail.com}
\end{itemize}

\begin{flushright}
\begin{tabular}{r}
%\includegraphics[width=0.2\linewidth]{src/2000px-Unihamburg-logo.png} % & \hfill & \includegraphics[width=0.4\linewidth]{logo.png}
\end{tabular}
\end{flushright}

>>>>

%\usepackage{wrapfig}
%\usepackage{fancyhdr}
\usepackage{float}
 %\usepackage{graphicx}
%\usepackage{subfig}
\usepackage{tikz}
\usetikzlibrary{spy}

\usepackage[position=t,singlelinecheck=off]{subfig}
\usepackage{caption}

\renewcommand\thesubfigure{\alph{subfigure}}

@@

\newlength{\sepwid}
\newlength{\onecolwid}
\newlength{\twocolwid}
\newlength{\threecolwid}
\setlength{\paperwidth}{1189mm} % A0 width: 46.8in
\setlength{\paperheight}{841mm} % A0 height: 33.1in
\setlength{\sepwid}{0.005\paperwidth} % Separation width (white space) between columns
\setlength{\onecolwid}{0.22\paperwidth} % Width of one column
\setlength{\twocolwid}{0.464\paperwidth} % Width of two columns
\setlength{\threecolwid}{0.708\paperwidth} % Width of three columns
\setlength{\topmargin}{-0.5in} % Reduce the top margin size
%-----------------------------------------------------------

@@

%\author{Jan Fabian Schmid \href{mailto:2schmid@informatik.uni-hamburg.de}{\textit{{\normalsize 2schmid@informatik.uni-hamburg.de}}}, Julian Kunkel \href{mailto:juliankunkel@googlemail.com}{\textit{{\normalsize juliankunkel@googlemail.com}}}} % Author(s)

@@

%\item \href{mailto:2schmid@informatik.uni-hamburg.de}{2schmid@informatik.uni-hamburg.de}
%\item \href{mailto:juliankunkel@googlemail.com}{juliankunkel@googlemail.com}

@@

\setlength{\sepwid}{0.02\paperwidth} % Separation width (white space) between columns

@@

For single hard disk systems such a model can be derived analytical \cite{Ruemmler94anintroduction}, however for the complex storage system of a super computer these models become too difficult to configure \cite{DBLP:conf/npc/ZhangLZJC10}.
Therefore we tried to find a good predictor for I/O-performance using a machine learning approach with artificial neural networks (ANNs).
A hypothesis was then made: \textbf{The I/O-path has significant influence on the time needed for a file access.}\medskip
%In particular the access time for a file access is more dependent on the I/O-path used in the storage system than on the access size for example.

To test our hypothesis we used ANNs with different input information for the prediction of access times.
For the use of I/O-paths as input information for the ANNs we developed a method to compute an approximation of the different I/O-paths the storage system used during a benchmark-test.

@@

	Our measurements showed, as can be seen in figure \ref{exploration}, that the access times increase with access sizes.
	However a phenomenon that can be explained with the I/O-path model can be seen as well; \textbf{access times of measurements with equal parameter values are split into several groups of access times.}
	The used I/O-paths for the file accesses distinguishes these groups of measurements through a step in the magnitude of access time.
	%Different I/O-paths were used for file accesses in these groups, which led to a step in the magnitude of the access time.

	\begin{figure}
		%\flushleft
		\captionsetup[subfigure]{margin={-0.4cm,0cm}}
		\subfloat[Sequential access pattern]{
			%\includegraphics[width=.55\textwidth]{src/plot_SizeSorted_log_read_seq.png}
			\begin{tikzpicture}[spy using outlines={circle,black,magnification=2.2,size=4.5cm, connect spies}]
			\node {\pgfimage[interpolate=true,width=.53\textwidth]{src/plot_SizeSorted_log_read_seq.png}};
			%\draw (3.5,0.5) ellipse (1 and 1);
			\spy on (3.4,0.5) in node [left] at (14,1.25);
			\end{tikzpicture}
		}
		%\hfill
		\\
		\captionsetup[subfigure]{margin={-0.4cm,0cm}}
		\vspace*{0.25cm}	
		\hspace*{-7.365cm}		
		\subfloat[Random access pattern]{
			%\hspace*{-5cm}
			\includegraphics[width=.53\textwidth]{src/plot_SizeSorted_log_read_rnd.png}
		}
		\vspace*{-0.05cm}	
		\caption{Access times of measured file accesses. The access sizes increase from left to right, all measurements with equal parameter values have the same color. In the highlighted section the described step in the access time occurs.}
		\label{exploration}
	\end{figure}

%	\begin{tikzpicture}[spy using outlines={circle,black,magnification=5,size=1.5cm, connect spies}]
%		\node {\pgfimage[width=.55\textwidth]{src/plot_SizeSorted_log_read_seq.png}};
%		\spy on (0,1) in node [left] at (10,1.25);
%	\end{tikzpicture}

	%\begin{tikzpicture}[spy using outlines={circle,black,magnification=2.5,size=4.5cm, connect spies}]
	%	\node {\pgfimage[interpolate=true,width=.55\textwidth]{src/plot_SizeSorted_log_read_seq.png}};
	%	\spy on (3.5,0.5) in node [left] at (14,1.25);
	%\end{tikzpicture}

@@

	Different models were used for the prediction of measured access times of file accesses.
	As baseline model linear regression was used for a simple mapping of access size to measured access time.
	Additionally three \textbf{models with different input information utilizing ANNs were used}.
	Every ANN-model received information about access sizes, file offsets and access types as input.
	One ANN also received information about the past data throughputs of the system, this can be used to exploit time dependencies of the I/O-performance in the system (see figure \ref{time_dep}). The last ANN received error classes as additional input.

	\begin{figure}
		\subfloat[In this sequence peaks in the access times occur regularly.]{
			\includegraphics[width=.48\textwidth]{src/plot_From100001to100250_read_seq.png}
		}
		\hfill
		\subfloat[Exponential moving average in red to predict peaks.]{
			\includegraphics[width=.48\textwidth]{src/ema2.png}
		}	
		\caption{Time dependencies can be utilized for better access time predictions.}
		\label{time_dep}
	\end{figure}

@@

The idea of error classes is based on the fact that \textbf{the residues of a simple model are characteristic for the groups of measurements with different typical access times} and therefore also for the I/O-paths used by the storage system for the corresponding file accesses. 
%Here a simple model is a model that doesn't discriminate measurements with similar parameters.
Error classes are then obtained by clustering the residues of such a model with a k-Means algorithm.

%	\begin{figure}
%		%\flushleft
%		%\subfloat[Sequential access pattern]{
%		%	\includegraphics[width=.55\textwidth]{src/linreg_error_clustering_seq_all.png}
%		%}
%		\begin{tikzpicture}[spy using outlines={circle,black}] %,magnification=2.5,size=4.5cm, connect spies}]
%		\node {\pgfimage[interpolate=true,width=.53\textwidth]{src/linreg_error_clustering_seq_all.png}};
%		\draw (3.7,0.9) ellipse (0.8 and 0.8);
%		%\spy on (3.5,0.5) in node [left]; % at (14,1.25);
%		\end{tikzpicture}
%		\\
%		%\hfill
%		\subfloat[Random access pattern]{
%			\includegraphics[width=.53\textwidth]{src/linreg_error_clustering_rnd_all.png}
%		}	
%		\caption{Residues from the linear regression model clustered into error classes. In the highlighted section the differen- tiation of measurements with varying I/O-paths can be seen.}
%		\label{error_classes}
%	\end{figure}

		\begin{figure}
			%\flushleft
			\captionsetup[subfigure]{margin={-0.4cm,0cm}}
			\subfloat[Sequential access pattern]{
				%\includegraphics[width=.55\textwidth]{src/plot_SizeSorted_log_read_seq.png}
				\begin{tikzpicture}[spy using outlines={circle,black,magnification=2.2,size=4cm, connect spies}]
				\node {\pgfimage[interpolate=true,width=.53\textwidth]{src/linreg_error_clustering_seq_all.png}};
				%\draw (3.5,0.5) ellipse (1 and 1);
				\spy on (3.5,0.9) in node [left] at (14,1.25);
				\end{tikzpicture}
			}
			%\hfill
			\\
			\captionsetup[subfigure]{margin={-0.4cm,0cm}}
			\hspace*{-7.365cm}	
			%\vspace*{0.25cm}	
			\subfloat[Random access pattern]{
				%\hspace*{-5cm}
				\includegraphics[width=.53\textwidth]{src/linreg_error_clustering_rnd_all.png}
			}
			\vspace*{-0.1cm}
			\caption{Residues from the linear regression model clustered into error classes. In the highlighted section the differen- tiation of measurements with varying I/O-paths can be seen.}
			\label{error_classes}
		\end{figure}

@@

	From the residues of the different models (see table \ref{residues}) it became clear that access times shouldn't be modelled linearly.
	Furthermore the exploitation of time dependencies of the storage system has been found to be difficult.
	As our hypothesis predicted \textbf{it is essential to use knowledge about I/O-paths for precise predictions of access times}.
	The error classes allow the ANN to decrease the error significantly.
	In figure \ref{preds} the difference in the predictions of the ANN without and with error class input can be seen.
	The second ANN is often able to classify file accesses correctly in respect to their I/O-path.

	\begin{table}
		\vspace{0ex}
		\begin{tabular}{l|r|r}%
			&  \multicolumn{2}{|c}{mean absolut error (s)}\\ \hline
			%\toprule
			Modell & seq. access & rnd. access\\ \hline
			%\midrule
			Linear regression & $7.6\cdot 10^-5$ & $4.76\cdot 10^-3$ \\
			ANN & $6.0\cdot 10^-5$ & $3.13\cdot 10^-3$ \\
			ANN + time dependencies & $5.7\cdot 10^-5$ & $3.05\cdot 10^-3$\\
			ANN + error classes & $2.0\cdot 10^-5$ & $1.03\cdot 10^-3$ \\
			%\bottomrule
		\end{tabular}
		\caption{Mean absolut errors of the different models for the two different file access patterns used in the benchmark-tests.}
		\label{residues}
	\end{table}

@@

	\vspace*{-0.75cm}
	\begin{figure}
		\subfloat[ANN without error classes]{
			\includegraphics[width=.47\textwidth]{src/plot_onlyPred_tuple1_Duration_rnd.png}
		}
		\hfill		
		\subfloat[ANN with error classes]{
			\includegraphics[width=.47\textwidth]{src/plot_onlyPred_tuple1_with_error_class_from_linreg_Duration_rnd.png}
		}	
		\caption{Measurements with random access pattern. Measured access times in black; predictions in blue.}
		\label{preds}
	\end{figure} 
%	
%	\begin{figure}
%		\includegraphics[width=0.48\linewidth]{src/plot_onlyPred_tuple1_Duration_rnd.png}
%		\hfill
%		\includegraphics[width=0.48\linewidth]{src/plot_onlyPred_tuple1_with_error_class_from_linreg_Duration_rnd.png}
%		\vspace*{-0.5cm}
%		\caption{Measurements in black; predictions in blue.}
%		\label{preds}
%	\end{figure}

@@

\nocite{*} % Insert publications even if they are not cited in the poster
\tiny{\bibliographystyle{unsrt}
\bibliography{sample}\vspace{0.75in}}

@@

	\footnotesize

	\setbeamercolor{block alerted title}{fg=black,bg=white} % Change the alert block title colors
	\setbeamercolor{block alerted body}{fg=black,bg=white} % Change the alert block body colors
	\vspace*{-2cm}
	\begin{exampleblock}{}
		\vspace*{-1.2cm}
		%\hspace*{5cm}
		\textbf{Contact Information:}
		\vspace*{-0.4cm}
		\begin{itemize}
			%\item Web: \href{http://wr.informatik.uni-hamburg.de}{http://wr.informatik.uni-hamburg.de}
			\item \href{mailto:2schmid@informatik.uni-hamburg.de}{2schmid@informatik.uni-hamburg.de}
			%\item \href{mailto:juliankunkel@googlemail.com}{juliankunkel@googlemail.com}
		\end{itemize}

	\end{exampleblock}

@@

%----------------------------------------------------------------------------------------
%	CONTACT INFORMATION
%----------------------------------------------------------------------------------------
%Apricot

%
%\begin{flushright}
%\begin{tabular}{r}
%%\includegraphics[width=0.2\linewidth]{src/2000px-Unihamburg-logo.png} % & \hfill & \includegraphics[width=0.4\linewidth]{logo.png}
%\end{tabular}
%\end{flushright}

### 2016-03-17 Thesis/bericht.tex
tabelle eingeführt, neue graphen, lupen, mehr referenzen

\subsubsection{Berechenbarkeitstheorie neuronaler Netze}
Vor der Anwendung von künstlichen neuronalen Netzen für ein Problem stellt sich die Frage, ob diese für das Problem geeignet sind. Dazu gibt es einige interessante mathematische Beweise.
Ein einzelnes Perzeptron kann alle linear separierbaren logischen Funktionen exakt approximieren (vgl. \cite{Rojas:1996:NNS:235222} S. 62-63), wobei lineare Separierbarkeit nach Rojas definiert ist als:

@@

\section{Leistungsvorhersage von Ein-/Ausgabe}
\label{rel_ea-vorhersage}
Das Problem, die Zugriffszeiten auf Festplatten vorherzusagen, kann im wesentlichen durch zwei verschiedene Ansätze gelöst werden (vgl. \cite{Crume:2013:FML:2538542.2538561} S. 45).
\begin{itemize}
\item  \textbf{White-Box-Modellierung:} Zum Einen kann man versuchen das Festplattensystem in einem Modell nachzubilden, indem Hardwaredetails, wie die Rotationsgeschwindigkeit der Platte, die Reaktionszeit und Geschwindigkeit des Lesekopfs, sowie das Zusammenspiel der Komponenten bekannt sind oder entsprechende Parameter durch gezielte Untersuchung approximiert werden. Mit diesem möglichst exakten Modell können dann Zugriffe simuliert und die Laufzeit des Modells gemessen werden. Diese Messung kann daraufhin als Vorhersage für das reale System verwendet werden. Eine White-Box ist ein System dessen innere Funktionsweise bekannt ist. Modelle dieses Ansatzes können entsprechend als White-Box-Modelle bezeichnet werden.
\item \textbf{Black-Box-Modellierung:} Beim zweiten Ansatz wird vom eigentlichen Festplattensystem abstrahiert und stattdessen ein mathematisches Modell gesucht, das das Verhalten des Systems möglichst genau beschreibt.
Zur Entwicklung des mathematischen Modells wird eine Regressionsanalyse durchgeführt. Dabei werden gemessene Leistungswerte von Festplattenzugriffen untersucht, um daraus passende Parameter für das Modell abzuleiten. Analog zur vorherigen Namensgebung kann dieser Ansatz als Black-Box-Modellierung bezeichnet werden, da hier ohne Wissen über die inneren Zustände des Systems modelliert wird.
\end{itemize}
Ich unterscheide daher im Folgenden zwischen der White-Box-Modellierung, bei der versucht wird das Festplattensystem nachzubilden (in der englischen Fachliteratur auch als \textit{analytic device modeling} und \textit{simulation modeling} bezeichnet) und der Black-Box-Modellierung, bei dem ein mathematisches Modell entwickelt wird.

>>>>

\subsubsection{Berechenbarkeitstheorie neuronaler Netze}
Vor der Anwendung von künstlichen neuronalen Netzen für ein Problem stellt sich die Frage, ob diese für das Problem geeignet sind. Dazu gibt es einige interessante mathematische Beweise.
Ein einzelnes Perzeptron kann alle linear separierbaren logischen Funktionen exakt approximieren (vgl. \
{Rojas:1996:NNS:235222} S. 62-63), wobei lineare Separierbarkeit nach Rojas definiert ist als:

@@

\section{Leistungsvorhersage von Ein-/Ausgabe}
\label{rel_ea-vorhersage}
Das Problem, die Zugriffszeiten auf Festplatten vorherzusagen, kann im wesentlichen durch zwei verschiedene Ansätze gelöst werden (vgl.{Crume:2013:FML:2538542.2538561} S. 45).
\begin{itemize}
\item  \textbf{White-Box-Modellierung:} Zum Einen kann man versuchen das Festplattensystem in einem Modell nachzubilden, indem Hardwaredetails, wie die Rotationsgeschwindigkeit der Platte, die Reaktionszeit und Geschwindigkeit des Lesekopfs, sowie das Zusammenspiel der Komponenten bekannt sind oder entsprechende Parameter durch gezielte Untersuchung approximiert werden. Mit diesem möglichst exakten Modell können dann Zugriffe simuliert und die Laufzeit des Modells gemessen werden. Diese Messung kann daraufhin als Vorhersage für das reale System verwendet werden. Eine White-Box ist ein System dessen innere Funktionsweise bekannt ist. Modelle dieses Ansatzes können entsprechend als White-Box-Modelle bezeichnet werden.
\item \textbf{Black-Box-Modellierung:} Beim zweiten Ansatz wird vom eigentlichen Festplattensystem abstrahiert und stattdessen ein mathematisches Modell gesucht, das das Verhalten des Systems möglichst genau beschreibt.
Zur Entwicklung des mathematischen Modells wird eine Regressionsanalyse durchgeführt. Dabei werden gemessene Leistungswerte von Festplattenzugriffen untersucht, um daraus passende Parameter für das Modell abzuleiten. Analog zur vorherigen Namensgebung kann dieser Ansatz als Black-Box-Modellierung bezeichnet werden, da hier ohne Wissen über die inneren Zustände des Systems modelliert wird.
\end{itemize}
Ich unterscheide daher im Folgenden zwischen der White-Box-Modellierung, bei der versucht wird das Festplattensystem nachzubilden (in der englischen Fachliteratur auch als \textit{analytic device modeling} und \textit{simulation modeling} bezeichnet) und der Black-Box-Modellierung, bei dem ein mathematisches Modell entwickelt wird.
